miniKanren with fair search strategies

### 2019-04-11 Paper/paper.tex
fix lots of grammar and spelling errors

Interleaving DFS, the search strategy of \conde in
miniKanren~\citep{friedman_reasoned_2018}, allocates computational resource
by the order in which the claueses are written. Each clause of a
\conde takes a half of the current resource and passes the other half
to its following clauses, except for the last clause that takes all of current
resource. 
both opportunity and burden: miniKanren users can place more frequently used 
goal at the beginning to optimize their programs; however, it might be a 
catastrophe if a goal that generate many useless states is placed before more 
important goals. Seasoned miniKanreners usually know how to utilize 
the unfairness to optimize their programs. However, we believe search strategies 
that is less sensitive to goal order can also be useful to little miniKanreners 
as well as seansoned ones. We propose two such search strategies, balanced 
interleaving DFS (biDFS) and breadth-first search (BFS), and observe how they 
affect the efficiency and the answer order of known miniKanren programs. The 
experiment is conducted with the miniKanren from \textit{The Reasoned Schemer, 
2nd Edition}.

@@

The above \texttt{run} generates 4 answers. All are lists of \texttt{*}s.
The order of the answers reflects the order miniKanren discovers them:
the first answer in the list is first discovered. This order is not suprising:
to generate the first answer, \texttt{'()}, miniKanren needs to apply
\texttt{repeato} only once and the later answers need more relation
applications. In this example, the cost of each answer is the same as
one more than the number of \texttt{*}s: the cost of \texttt{'()} is 1,
the cost of \texttt{'(*)} is 2, and so on.

@@

With iDFS, \texttt{'(a a)} occurs before \texttt{'(b)} while \texttt{'(a a)} is 
associated with a higher cost. iDFS strategy is the cause, since it prioritizes 
the first \conde case considerablely. When every \conde case 
are equally productive, the iDFS strategy takes $1/2^{i}$ answers from the 
$i$-th case, except the last case, which share the same portion as the second 
last one. In contrast, the same run with BFS produces answers in an expected 
order.

@@

Our first solution, balanced interleaving DFS (biDFS), is not fair as well. 
However, it is less sensitive to goal order in disjucntion and is as efficient 
as iDFS. This search strategy is based on the  observations that a disjuction 
can be viewed as a binary tree, where \texttt{disj2}s are nodes 
and sub-goals are leaves, and that the position of a leaf has direct relation 
with its priority -- the deeper a leaf, the lower computational effort it is 
shared. In iDFS, the tree is in one of the most unbalanced forms, because 
\texttt{disj} applys \texttt{disj2} right associatively. Hence the goal order 
is very important. The reason why the tree shape can determine priority is that 
\texttt{disj2} allocates computational effort evenly to its two sub-goals.

@@

Breadth-first search (BFS) is fair. Our BFS is similar to the one by Seres et 
al \citet{seres1999algebra}. We conjecture that the two strategies produce
answers in the same order. ;; need more comparision

In the first subsection, we change the search strategy from iDFS to BFS. In the 
second subsection we optimize the search engine by using a queue to manage the 
search order. The two subsections corresponds to two new version of miniKanren, 
\texttt{mk-1} and \texttt{mk-2}. We refer to the original version from TRS2 as 
\texttt{mk-0}.

@@

In both \texttt{mk-0} and \texttt{mk-1}, search spaces are represented by 
streams of  answers. The thunky streams in \texttt{mk-0} denote delayed 
computation, however, they do not necessary mean an increment in cost. We use 
the same kind of stream in \texttt{mk-1} but only put thunk at those places 
where the cost of following answers is increased by one.

To ease discussion, we call the \texttt{car}s of an stream its \emph{mature} 
part, and the last \texttt{cdr} the \emph{immature} part. When the stream is 
definitely finite, its immature part is an empty list, otherwise, it is a 
thunk. We sometimes say a stream is immature to mean its mature part is empty.

Streams denote cost correctly when they are constructed by \texttt{==}, 
\texttt{succeed}, and \texttt{fail}. However, the \texttt{mk-0} version of 
\texttt{append-inf} (Fig.~\ref{append-inf-0}) breaks cost respectiveness if its 
first input stream, \texttt{s-inf},  is infinite. The resulting mature part 
contains only the mature part of \texttt{s-inf}. If we want to encode the 
cost information correctly, the resulting mature part should also contains the 
mature part of \texttt{t-inf}.

@@

The \texttt{mk-1} version of \texttt{append-inf} (Fig.~\ref{append-inf-1}) 
gain fairness by combining the mature parts in the fashion of \texttt{append}. 
This \texttt{append-inf} calls its helper immediately, with the first argument, 
\texttt{s?}, set to \texttt{\#{}t}, which means \texttt{s-inf} in the helper is 
the \texttt{s-inf} in the driver. Two streams are swapped in the third 
\texttt{cond} clause, with \texttt{s?} flipped accordingly.

@@

\texttt{mk-1} is not efficient in two aspects. \texttt{append-inf} need to copy 
all \texttt{cons} cells of \emph{both} input streams when the first stream is 
possibly infinite. Besides, \texttt{mk-1} computes answers of the same cost
at once, even when only a portion is queried. We solves the two problems in the 
next subsections.

@@

We avoid generating same-cost answers at once by expressing BFS with a queue, 
whose elements thunks that return a new stream. Every \texttt{mk-1} stream has 
zero or one thunk, so it is uninteresting to manage them with queue . Therefore 
we change the representation of immature parts from thunks to 
lists of thunks. As as consequence, we also change the way to combine mature and 
immature part from \texttt{append} to \texttt{cons}.

@@

The first three functions are constructors: \texttt{empty-inf} constructs an 
empty stream; \texttt{unit-mature-inf} constructs a stream with one mature 
solution; \texttt{unit-immature-inf} constructs a stream with one thunk.

The \texttt{append-inf} in \texttt{mk-3} is relatively straightforwared compared 
with the \texttt{mk-1} version. \texttt{append-map-inf} is more tricky on how to 
construct the new immature part. We can follow the approach in mk-0 and mk-1 -- 
create a new thunk which invoke append-map-inf recursively when forced. But then 
we need to be careful: if we construct the thunk when the old immature part is 
an empty list, the resulting stream might be infinitely unproductive. Beside, 
all solutions of the next lowest cost in \texttt{s-inf} must be computed when 
the thunk is invoked. However sometimes only a portion of these solutions is 
required to answer a query. To avoid the trouble and the too-early computation, 
we choose to create a new thunk for every existing thunk.

The next four functions are used only by \texttt{ifte} and \texttt{once}. 
Uninterested readers might skip them. \texttt{null-inf?} checks whether a stream 
is exausted. \texttt{mature-inf?} checks whether a stream has some mature 
solutions. \texttt{car-inf} takes the first solution out of a mature stream. 
\texttt{cdr-inf} drops the first solution of a mature stream. Finally, 
\texttt{force-inf} forces an a immature stream to do more computation.

\texttt{unit-mature-inf}, \texttt{append-map-inf}, \texttt{empty-inf} and 
\texttt{append-inf} form a \emph{MonadPlus}, where they corresponds to 
\texttt{unit}, \texttt{bind}, \texttt{mzero}, and \texttt{mplus} respectively.

@@

The last interesting function is \texttt{take-inf} (Fig.~\ref{take-inf-3-1}). 
The parameter \texttt{vs} is a list of solutions. The next two parameters, 
\texttt{P} and \texttt{Q}, together represent a queue. The first two 
\texttt{cond} lines are very similar to their counterparts in mk-0 and mk-1. 
The third line runs when we exaust all solutions. The forth line re-shape the 
queue. The fifth and last line invoke the first thunk in the queue and use the 
mature part of the resulting stream, \texttt{s-inf}, as the new \texttt{vs}, 
and enqueuing \texttt{s-inf}'s thunks.

@@

We also devise breadth-first search, which is similar to Seres's BFS. These two 
BFSs have similar intuition, but whether they are equivalent is not clear yet. 
Our optimization on BFS with queue, however, is new.

>>>>

Interleaving DFS, the search strategy of \conde in
miniKanren~\citep{friedman_reasoned_2018}, allocates computational resource
by the order in which the clauses are written. Each clause of a
\conde takes half of the current resource and passes the other half
to its following clauses, except for the last clause that takes all of current
resource. 
both opportunity and burden: miniKanren users can place more frequently used 
goal at the beginning to optimize their programs; however, it might be a 
catastrophe if a goal that generate many useless states is placed before more 
important goals. Seasoned miniKanreners usually know how to utilize 
the unfairness to optimize their programs. However, we believe search strategies 
that is less sensitive to goal order can also be useful to little miniKanreners 
as well as seasoned ones. We propose two such search strategies, balanced 
interleaving DFS (biDFS) and breadth-first search (BFS), and observe how they 
affect the efficiency and the answer order of known miniKanren programs. The 
experiment is conducted with the miniKanren from \textit{The Reasoned Schemer, 
2nd Edition}.

@@

The above \texttt{run} generates 4 answers. All are lists of \texttt{*}s.
The order of the answers reflects the order miniKanren discovers them:
the first answer in the list is first discovered. This order is not surprising:
to generate the first answer, \texttt{'()}, miniKanren needs to apply
\texttt{repeato} only once and the later answers need more relation
applications. In this example, the cost of each answer is the same as
one more than the number of \texttt{*}s: the cost of \texttt{'()} is 1,
the cost of \texttt{'(*)} is 2, and so on.

@@

With iDFS, \texttt{'(a a)} occurs before \texttt{'(b)} while \texttt{'(a a)} is 
associated with a higher cost. iDFS strategy is the cause, since it prioritizes 
the first \conde case considerably. When every \conde case 
are equally productive, the iDFS strategy takes $1/2^{i}$ answers from the 
$i$-th case, except the last case, which share the same portion as the second 
last one. In contrast, the same run with BFS produces answers in an expected 
order.

@@

Our first solution, balanced interleaving DFS (biDFS), is not fair as well. 
However, it is less sensitive to goal order in disjucntion and is as efficient 
as iDFS. This search strategy is based on the  observations that a disjuction 
can be viewed as a binary tree, where \texttt{disj2}s are nodes 
and sub-goals are leaves, and that the position of a leaf has direct relation 
with its priority -- the deeper a leaf, the lower computational effort it is 
shared. In iDFS, the tree is in one of the most unbalanced forms, because 
\texttt{disj} applys \texttt{disj2} right associatively. Hence, the goal order 
is very important. The reason why the tree shape can determine priority is that 
\texttt{disj2} allocates computational effort evenly to its two sub-goals.

@@

Breadth-first search (BFS) is fair. Our BFS is similar to the one by Seres et 
al \citet{seres1999algebra}. We conjecture that the two strategies produce
answers in the same order.
%TODO need more comparison

In the first subsection, we change the search strategy from iDFS to BFS. In the 
second subsection we optimize the search engine by using a queue to manage the 
search order. The two subsections corresponds to two new versions of miniKanren, 
\texttt{mk-1} and \texttt{mk-2}. We refer to the original version from TRS2 as 
\texttt{mk-0}.

@@

In both \texttt{mk-0} and \texttt{mk-1}, search spaces are represented by 
streams of answers. Thunk streams in \texttt{mk-0} denote delayed 
computation, however, they do not necessary mean an increment in cost. We use 
the same kind of stream in \texttt{mk-1} but only put thunk at those places 
where the cost of following answers is increased by one.

To ease discussion, we call the \texttt{car}s of a stream as its \emph{mature} 
part, and its last \texttt{cdr} as its \emph{immature} part. When the stream is 
definitely finite, its immature part is an empty list, otherwise, it is a 
thunk. We sometimes say a stream is immature to mean its mature part is empty.

Streams denote cost correctly when they are constructed by \texttt{==}, 
\texttt{succeed}, and \texttt{fail}. However, the \texttt{mk-0} version of 
\texttt{append-inf} (Fig.~\ref{append-inf-0}) breaks cost respectiveness if its 
first input stream, \texttt{s-inf},  is infinite. The resulting mature part 
contains only the mature part of \texttt{s-inf}. If we want to encode the 
cost information correctly, the resulting mature part should also contain the 
mature part of \texttt{t-inf}.

@@

The \texttt{mk-1} version of \texttt{append-inf} (Fig.~\ref{append-inf-1}) 
gain fairness by combining the mature parts in the fashion of \texttt{append}. 
This \texttt{append-inf} calls its helper immediately, with the first argument, 
\texttt{s?}, set to \texttt{\#{}t}, which \texttt{s-inf} and \texttt{t-inf} haven't been swapped in the driver. They are swapped in the third \texttt{cond} clause, where \texttt{s?} is flipped accordingly.

@@

\texttt{mk-1} is not efficient in two aspects. \texttt{append-inf} need to copy 
all \texttt{cons} cells of \emph{both} input streams when the first stream is 
possibly infinite. Besides, \texttt{mk-1} computes answers of the same cost
at once, even when only a portion is queried. We solve the two problems in the 
next subsections.

@@

We avoid generating same-cost answers at once by expressing BFS with a queue, 
whose elements thunks that return a new stream. Every \texttt{mk-1} stream has 
zero or one thunk, so it is uninteresting to manage them with queue . Therefore, 
we change the representation of immature parts from thunks to 
lists of thunks. Consequently, we also change the way to combine mature and 
immature part from \texttt{append} to \texttt{cons}.

@@

The first three functions are constructors: \texttt{empty-inf} makes an 
empty stream; \texttt{unit-mature-inf} makes a stream with one mature 
solution; \texttt{unit-immature-inf} makes a stream with one thunk.

The \texttt{append-inf} in \texttt{mk-3} is relatively straightforward compared 
with the \texttt{mk-1} version. \texttt{append-map-inf} is more tricky on how to 
construct the new immature part. We can follow the approach in mk-0 and mk-1 -- 
create a new thunk which invoke append-map-inf recursively when forced. But then 
we need to be careful: if we construct the thunk when the old immature part is 
an empty list, the resulting stream might be infinitely unproductive. Beside, 
all solutions of the next lowest cost in \texttt{s-inf} must be computed when 
the thunk is invoked. However, sometimes only a portion of these solutions is 
required to answer a query. To avoid the trouble and too-early computation, 
we choose to create a new thunk for every existing thunk.

The next four functions are used only by \texttt{ifte} and \texttt{once}. 
Uninterested readers might skip them. \texttt{null-inf?} checks whether a stream 
is exhausted. \texttt{mature-inf?} checks whether a stream has some mature 
solutions. \texttt{car-inf} takes the first solution out of a mature stream. 
\texttt{cdr-inf} drops the first solution of a mature stream. Finally, 
\texttt{force-inf} forces an immature stream to do more computation.

\texttt{unit-mature-inf}, \texttt{append-map-inf}, \texttt{empty-inf} and 
\texttt{append-inf} form a \emph{MonadPlus}, where they correspond to 
\texttt{unit}, \texttt{bind}, \texttt{mzero}, and \texttt{mplus} respectively.

@@

The last interesting function is \texttt{take-inf} (Fig.~\ref{take-inf-3-1}). 
Its first parameter, \texttt{vs}, is a list of solutions. The next two parameters, \texttt{P} and \texttt{Q}, together represents a queue. 
The first two \texttt{cond} lines are very similar to their counterparts in mk-0 and mk-1. The third line runs when both the answer list \texttt{vs} and the queue are empty, which means we have found all answers. The forth line re-shape the queue. The fifth and last line invoke the first thunk in the queue and use the mature part of the resulting stream, \texttt{s-inf}, as the new \texttt{vs}, and enqueuing \texttt{s-inf}'s thunks.

@@

We also devise breadth-first search, which is similar to Seres's BFS. It is not clear whether they are equivalent. Our optimization on BFS with queue, however, is new.
